The Pair Distribution Function of the Inhomogeneous Hard-Sphere Fluid

### 2014-03-06 papers/pair-correlation/paper.tex

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We have developed an approximation for the pair distribution
  function of hard spheres, which is based on the average value of the
  pair distribution function at contact described in
  Section~\ref{sec:contact}.  Our pair distribution function is constructed so
  as to be computable with only single-site convolutions and to
  accurately reproduce the averaged distribution function at contact.
  This results in a functional that is efficient to compute, and
  predicts the pair distribution function as well as the most accurate
  existing analytical approximation.
\end{abstract}

@@

The canonical inhomogeneous configuration for the hard-sphere
fluid is the system consisting of a hard sphere at a hard wall.  In
1986, Plischke and Henderson solved the pair distribution function of
this system using integral equation theory under the Percus-Yevick
approximation~\cite{plischke1986pair}.  Ten years later,
G{\"o}tzelmann \emph{et al.} found the pair distribution function near
a hard wall using classical DFT by finding the direct correlation
function by taking a functional derivative of the free energy
functional and then solving the Ornstein-Zernicke
equation~\cite{gotzelmann1996structure}.  The approach of solving the
Ornstein-Zernicke equation for an inhomogeneous system is
computationally challenging~\cite{paul2003variational}.

@@

\section{Approximations to the pair distribution function}
There are few analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  These works differ in what density they
use when evaluating the radial distribution function $g(r,\eta)$.

Several recent works have cited a
mean-density approximation:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \frac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter~\cite{gloor2007prediction, gross2009density}.  These works
did not however use this approach, which fails dramatically at higher
densities, as the packing fraction found from the mean density exceeds
unity.

Non-divergent approaches use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
where $\sigma$ is the hard-sphere diameter.  Equation~\ref{eq:fischer}
is computationally awkward, because it relies on the midpoint
$\tfrac12(\rr_1+\rr_2)$.  Moreover, the approach of Fischer and
Methfessel is intended to approximate the pair distribution function only
at contact, when the distance between $\rr_1$ and $\rr_2$ is the hard-sphere
diameter.

@@

\section{Our approach}
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
evaluated at the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
\end{align}

In the approaches for the pair distribution function mentioned above,
function forms of radial distributions (used to approximate pair
correlation) were dependent upon independent variables $\eta$, the
density, whether at single points or averaged over sourrounding
volumes.  We can achieve greater accuracy by using instead a function
dependent upon our $g_{\sigma}$, which holds more information about an
inhomogeneous system then a simple convolution of the density does.
We therefore seek a function form for our radial distribution that is
dependent upon $g_{\sigma}$ and that will approach the correct
solution in the homogeneous limit and also in the limit of contact
distance between the two points.

@@

The function has the form:
\davidsays{fix below to match next section... or should we just delete it,
and express the concept in words? It seems maybe redundant.}
\begin{multline}
  g(|\rr|) = 1 + h_{\sigma}e^{-\kappa_0\left(\frac{r}{\sigma}-1\right)} \\+ x_1h_{\sigma}\sin(x_2(r-\sigma))e^{-x_3(r-\sigma)}\\
  -x_4h_{\sigma}^2\sin(x_5(r-\sigma))e^{-x_6(r-\sigma)}
\end{multline}
The finalized form of our pair correlation function is:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
\end{align}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}

@@

We construct $g(r, g_\sigma)$ to satisfy several constraints.  Naturally,
\begin{equation}
  g(\sigma, g_\sigma) = g_\sigma.
\end{equation}
A second constraint applies to the integral of the total correlation
function
\begin{align}
  1 + n\int h(r)d\rr &= nkT\chi_T \label{eq:total-constraint}
\end{align}
where $h(r) = g(r) - 1$ is the total correlation function.
A final constraint is that the slope of the radial distribution
function at contact is given to a very good approximation by
\begin{align}
  g'(\sigma) \approx - h_\sigma g_\sigma
\end{align}
where $h_\sigma = g_\sigma - 1$ is the total correlation function.
Under these three constraints, we applied a least-squares fit to
simulation data for the radial distribution function at packing
fractions from 0.05 to 0.4 by steps of 0.05.  Our fitted radial
distribution function is given by
\newcommand\kappaO{\kappa_0}
\newcommand\kappaI{\kappa_1}
\newcommand\kappaZ{\kappa_2}
\begin{multline}
  g(r,g_\sigma) = 1 + h_\sigma e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + h_\sigma(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
\begin{equation}
  I(g_\sigma) = \kappaZ^5 \frac{
    \frac{\chi-1}{24\eta} - h_\sigma \frac{(\kappaO-2
      g_\sigma)(\kappaI^2 + 4 \kappaI + 6)}{\kappaI^4}
    - h_\sigma\frac{\kappaO^2 + 2\kappaO + 2}{\kappaO}
  }{2 \kappaZ^2 + 12 \kappaZ + 24}
\end{equation}
\begin{equation}
  \chi = \frac{(1-\eta)^4}{\eta^4 - 4\eta^3 + 4\eta^2 + 4\eta + 1}
\end{equation}
\newcommand\nastyetacuberoot{\sqrt[3]{54 g_\sigma^2 -
    6\sqrt{81g_\sigma^4 - 6g_\sigma^3}}}
\begin{equation}
  \eta = 1 - \frac{1}{\nastyetacuberoot} - \frac{\nastyetacuberoot}{6g_\sigma}
\end{equation}
This function includes three fitted parameters, $\kappaO =
\kappazero$, $\kappaI = \kappaone$,
and $\kappaZ = \kappatwo$.
The fit is displayed in Fig.~\ref{fig:radial-distribution}.  In that
figure, it is apparent that the integral f $h(r)$ from
Eq.~\ref{eq:total-constraint} results in cancelling errors at large
distances.  The maximum error in $g(r)$ at packing fractions up to
0.4 is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr$.

>>>>

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We have developed an approximation for the pair distribution
  function of hard spheres, which is based on the average value of the
  pair distribution function at contact we recently developed.  Our
  pair distribution function is constructed so as to be computable
  with only single-site convolutions and to accurately reproduce the
  averaged distribution function at contact.  This results in a
  functional that is efficient to compute, and predicts the pair
  distribution function as well as the most accurate existing
  analytical approximation.  \fixme{needs special attention.}
\end{abstract}

@@

The canonical inhomogeneous configuration for the hard-sphere fluid is
the system consisting of a hard sphere at a hard wall.  In 1986,
Plischke and Henderson solved the pair distribution function of this
system using integral equation theory under the Percus-Yevick
approximation~\cite{plischke1986pair}.  Ten years later,
G{\"o}tzelmann \emph{et al.} found the pair distribution function near
a hard wall by solving the Ornstein-Zernicke equation using the direct
correlation function found from a functional derivative of a classical
DFT free energy functional~\cite{gotzelmann1996structure}.
Unfortunately, solving the inhomogeneous Ornstein-Zernicke equation is
computationally challenging, although more efficient approximate
algorithms have been developed~\cite{paul2003variational}.

@@

\section{Approximations to the pair distribution function}
There are several analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  These works differ in what density is
used when evaluating the radial distribution function $g(r,\eta)$.

Several recent works have assumed a mean-density approximation for
the pair distribution function:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter~\cite{gloor2007prediction, gross2009density}.  This approach,
however, cannot be applied to highly inhomogeneous systems such as a
dense fluid at a solid surface.  These systems exhibit large density
oscillations that lead to mean packing fractions greater than unity,
which cannot occur in the bulk reference system that defines $g(r,
\eta)$.  The above papers restricted themselves to the liquid-vapor
interface, which does not exhibit this pathology.

Non-pathological approaches use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
where $\sigma$ is the hard-sphere diameter.  Equation~\ref{eq:fischer}
is computationally awkward, because it treats as special the midpoint
$\tfrac12(\rr_1+\rr_2)$.  Moreover, the approach of Fischer and
Methfessel is intended to approximate the pair distribution function only
at contact, when the distance between $\rr_1$ and $\rr_2$ is the hard-sphere
diameter.

@@

\section{Our approach}
%% We approximate the pair correlation with the average of two radial
%% distribution functions, evaluated at the distance between the two
%% points, that are themselves functions of the averaged pair correlation
%% at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
%% evaluated at the two points.
In the approaches for the pair distribution function mentioned above,
the radial distribution function used to approximate pair correlation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by using instead a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}.
\end{align}

@@

\newcommand\kappaO{\kappa_0}
\newcommand\kappaI{\kappa_1}
\newcommand\kappaZ{\kappa_2}

The function has the form:
\begin{multline}
  g(r,g_\sigma) = 1 + h_\sigma e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + h_\sigma(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
where $\kappaO$,$\kappaI$, and $\kappaZ$ are fitted constants and
$I(g_\sigma)$ is a function of $g_\sigma$.

The finalized form of our pair correlation function is:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
\end{align}
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}

@@

We construct $g(r, g_\sigma)$ to satisfy several constraints.  Naturally,
\begin{equation}
  g(\sigma, g_\sigma) = g_\sigma.
\end{equation}
A second constraint applies to the integral of the total correlation
function
\begin{align}
  1 + n\int h(r)d\rr &= nkT\chi_T \label{eq:total-constraint}
\end{align}
where $h(r) = g(r) - 1$ is the total correlation function.
A final constraint is that the slope of the radial distribution
function at contact is given to a very good approximation by
\begin{align}
  g'(\sigma) \approx - h_\sigma g_\sigma
\end{align}
where $h_\sigma = g_\sigma - 1$ is the total correlation function.
Under these three constraints, we applied a least-squares fit to
simulation data for the radial distribution function at packing
fractions from 0.05 to 0.4 by steps of 0.05.  Our fitted radial
distribution function is given by
\begin{multline}
  g(r,g_\sigma) = 1 + h_\sigma e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + h_\sigma(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
\begin{equation}
  I(g_\sigma) = \kappaZ^5 \frac{
    \frac{\chi-1}{24\eta} - h_\sigma \frac{(\kappaO-2
      g_\sigma)(\kappaI^2 + 4 \kappaI + 6)}{\kappaI^4}
    - h_\sigma\frac{\kappaO^2 + 2\kappaO + 2}{\kappaO}
  }{2 \kappaZ^2 + 12 \kappaZ + 24}
\end{equation}
\begin{equation}
  \chi = \frac{(1-\eta)^4}{\eta^4 - 4\eta^3 + 4\eta^2 + 4\eta + 1}
\end{equation}
\newcommand\nastyetacuberoot{\sqrt[3]{54 g_\sigma^2 -
    6\sqrt{81g_\sigma^4 - 6g_\sigma^3}}}
\begin{equation}
  \eta = 1 - \frac{1}{\nastyetacuberoot} - \frac{\nastyetacuberoot}{6g_\sigma}
\end{equation}
This function includes three fitted parameters, $\kappaO =
\kappazero$, $\kappaI = \kappaone$,
and $\kappaZ = \kappatwo$.
The fit is displayed in Fig.~\ref{fig:radial-distribution}.  In that
figure, it is apparent that the integral f $h(r)$ from
Eq.~\ref{eq:total-constraint} results in cancelling errors at large
distances.  The maximum error in $g(r)$ at packing fractions up to
0.4 is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr$.
